Lesson 4
========================================================

***

```{r}
library(ggplot2) #must load the ggplot package first
data(diamonds) #loads the diamonds data set since it comes with the ggplot package
summary(diamonds)
str(diamonds)
levels(diamonds$color)
?diamonds
```

***

```{r}
qplot(
  data = diamonds,
  x = price
)
```

***

```{r}
summary(diamonds$price)
```

```{r}
str(subset(diamonds, price >= 15000)$price)
```

***

```{r}
qplot(
  data = diamonds,
  x = price,
  binwidth = 20
) +
  scale_x_continuous(breaks = seq(300, 1600, 50), limits = c(300, 1600))
```

***

```{r}
qplot(
  data = diamonds,
  x = price
) +
  facet_wrap(~cut, ncol = 1, scales = "free_y")
?facet_wrap
```

```{r}
by(diamonds$price, diamonds$cut, summary)
```

***

```{r}
qplot(
  data = diamonds,
  x = price / carat,
  binwidth = 0.03
) +
  facet_wrap(~cut, ncol = 1, scales = "free_y") +
  scale_x_log10()
```

```{r}
names(diamonds)
boxplot(
  price~color,
  data = diamonds
)
```

***

```{r}
by(diamonds$price, diamonds$color, summary)
```

```{r}
7695 - 1860
```

```{r}
4214 - 911
```

***

```{r}
boxplot(
  price~color,
  data = diamonds
)
```

***

```{r}
qplot(
  data = diamonds,
  x = carat,
  binwidth = 0.01
) +
  scale_x_continuous(breaks = seq(0, 2, 0.1), limits = c(0, 2))
```

***

### Scatterplots and Perceived Audience Size
Notes:

*****

```{r}
install.packages(c("dplyr", "tidyr"))
library(dplyr)
library(tidyr)
```

```{r}
mm <- read.csv(
  'murdered male indicator file UL 20100916 - Data.csv',
  header = T,
  row.names = 1,
  check.names = F)
```

```{r}
mm[1:2,1:2]
str(mm)
rownames(mm)
mmc <- mm %>% 
  mutate(country = rownames(mm)) %>% 
  gather('year', 'pp', 1:55)
```

```{r}
arrange(mmc, desc(pp))

mts <- mmc %>% facet_wrap(~country) %>% ts(start=c(1950), end=c(2004), frequency=1)
plot(mts)

mmc %>% ggplot(aes(x=year, y=pp), color=country) +    geom_line()
    # facet_wrap(~country, ncol = 1)
```

*****

```{r}
birthdays <- read.csv('birthdaysExample.csv')


qplot(
  data = birthdays,
  x = as.Date(dates, '%m/%d/%y'),
  binwidth = 7 # = 60 * 60 * 24 * 7
) # + scale_x_date(breaks = seq(0, 365, 1))
```

***

```{r}
library(ggplot2)

pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
```

### Scatterplots
Notes:

```{r Scatterplots}
qplot(
  data = pf,
  x = age,
  y = friend_count
)
```

***

#### What are some things that you notice right away?
Response:

***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
ggplot(
  aes(x = age, y = friend_count),
  data = pf
) +
  geom_point() +
  xlim(13, 90)
```

***

### Overplotting
Notes:

```{r Overplotting}
ggplot(
  aes(x = age, y = friend_count),
  data = pf
) +
  geom_point(alpha = 1 / 20) +
  xlim(13, 90)
```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:

```{r Coord_trans()}
ggplot(
  aes(x = age, y = friend_count),
  data = pf
) +
  geom_point(alpha = 1 / 20) +
  xlim(13, 90) +
  coord_trans(y = 'sqrt')
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}
ggplot(
  aes(y = friendships_initiated, x = age),
  data = pf
) +
  geom_point(alpha = 1 / 20) +
  xlim(13, 90) +
  coord_trans(y = 'sqrt')
```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
ggplot(
  aes(x = age, y = friend_count),
  data = pf
) +
  geom_jitter(alpha = 1 / 20) +
  xlim(13, 90)
```

***

```{r}
library(dplyr)
```

```{r}
pf.summary.by.age <- pf %>%
  group_by(age) %>%
  summarise(
    friend.count.mean = mean(friend_count),
    friend.count.median = median(friend_count),
    n = n()
  )
```

```{r}
ggplot(
  data = pf.summary.by.age,
  aes(x = age, y = friend.count.mean)
) +
  geom_line() +
  xlim(12, 90)
```

***

```{r}
ggplot(
  data = pf,
  aes(x = age, y = friend_count)
) +
  geom_point(alpha = 0.05, color = 'orange') +
  xlim(13, 90) +
  coord_trans(y = 'sqrt') +
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = 0.1), linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = 0.9), linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = 0.5), color = 'blue')
```

```{r}
ggplot(
  data = pf,
  aes(x = age, y = friend_count)
) +
  geom_point(alpha = 0.05, color = 'orange') +
  coord_cartesian(xlim = c(13, 70), ylim = c(0, 1000)) +
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = 0.1), linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = 0.9), linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = 0.5), color = 'blue')
```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:

```{r Conditional Means}

```

Create your plot!

```{r Conditional Means Plot}

```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}

```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
cor.test(
  x = pf$age,
  y = pf$friend_count,
  method = "pearson",
  alternative = "two.sided"
)
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(subset(pf, age <= 70), cor.test(age, friend_count))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
ggplot(
  data = pf,
  aes(x = likes_received, y = www_likes_received)
) +
  geom_point(alpha = 0.05) +
  coord_cartesian(
    xlim = c(0, quantile(pf$likes_received, 0.95)),
    ylim = c(0, quantile(pf$www_likes_received, 0.95))
  )
```

***

### Strong Correlations
Notes:

```{r Strong Correlations}

```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
cor.test(
  x = pf$www_likes_received,
  y = pf$likes_received
)
```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
install.packages('alr3')
library(alr3)
data("Mitchell")
?Mitchell
head(Mitchell)
```

Create your plot!

```{r Temp vs Month}
ggplot(
  data = Mitchell,
  aes(x = Month, y = Temp)
) +
  geom_point(alpha = 0.5)
```

```{r}
cor.test(
  x = Mitchell$Month,
  y = Mitchell$Temp
)
```

```{r}
###: TODO wrap plot ###
ggplot(
  data = Mitchell,
  aes(
    x = Month %% 12,
    y = Temp
  )
) +
  geom_point(alpha = 0.5)

```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r}
pf$age_with_months <- pf$age + (1 - pf$dob_month / 12)
```

```{r Noisy Scatterplots}
ggplot(
  data = pf,
  aes(x = age_with_months, y = friend_count)
) +
  geom_line(stat = 'summary', fun.y = mean)
```

```{r}
pf.fc_by_age_months <- pf %>%
  arrange(age_with_months) %>%
  group_by(age_with_months) %>%
  summarise(
    friend_count_mean = mean(friend_count),
    friend_count_median = median(friend_count),
    n = n()
  )
```

```{r}
ggplot(
  data = pf.fc_by_age_months,
  aes(x = age_with_months, y = friend_count_mean)
) +
  geom_line(color = 'orange') +
  geom_smooth() +
  xlim(13, 71) +
  ylim(0, 450)
```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}

```

***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}

```

***

### Age with Months Means

```{r Age with Months Means}

```

Programming Assignment
```{r Programming Assignment}

```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}

```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}

```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

