
Loan Data from Prosper Exploration by Justinas Marozas
======================================================

This report explores loan data from Prosper with 113937 observations and 81 
data points extracted. As of time of writing, brief descriptions of data 
points can be found [here](https://docs.google.com/spreadsheets/d/1gDyi_L4UvIrLTEC6Wri5nbaMmkGmLQBk-Yx3z0XDEtI/edit#gid=0).


## Changes to the dataset and considerations

- Loans as a domain is largely unknown to me. With 81 odd datapoint to figure 
out this report will be more of a lookaround than a focussed analysis;
- An observation in the dataset is a snapshot of a loan listing. Time series 
analysis of a listing is not an option, but grouping by MemberKey it would be 
possible to run time series analysis on borrowers. With 90831 unique values 
of MemberKey there's few members that would have enough listings to be worth 
the effort;
- I'll convert ListingCreationDate, ClosedDate, DateCreditPulled, 
LoanOriginationDate, FirstRecordedCreditLine do datetime type;
- I'll convert LoanOriginationQuarter to a chronologically ordered factor;
- I'll drop "Not Employed" and "Not Displayed" levels from IncomeRange and 
order it. It would be better to create a separate datapoint for this because 
actual income ranges and two dropped levels all have distinct meanings, but 
not particularly interested in that at the moment;
- I'll convert ProsperRating..Alpha. and CreditGrade into ordered factors. 
I'll also introduce data point CreditGrade.ProsperRating that is a joined 
version of ProsperRating..Alpha. and CreditGrade, because their levels seem to 
carry same meaning and a listing never has both present;
- I'll introduce data point ListingCategory based on ListingCategory..numeric. 
and description in the data point definition;
- I'll convert IncomeVerifiable, CurrentlyInGroup, IsBorrowerHomeowner to \
boolean type;
- Ill introduce past.due.days ordered factor based on LoanStatus.


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(dplyr)
library(stringr)
library(lubridate)
library(RColorBrewer)
```

```{r echo=FALSE, Load_the_DataSet}
loans <- read.csv('prosperLoanData.csv')
```

```{r echo=FALSE, Tidy_up}

# The category of the listing that the borrower selected when posting their 
# listing: 0 - Not Available, 1 - Debt Consolidation, 2 - Home Improvement, 
# 3 - Business, 4 - Personal Loan, 5 - Student Use, 6 - Auto, 7- Other, 
# 8 - Baby&Adoption, 9 - Boat, 10 - Cosmetic Procedure, 11 - Engagement Ring, 
# 12 - Green Loans, 13 - Household Expenses, 14 - Large Purchases, 
# 15 - Medical/Dental, 16 - Motorcycle, 17 - RV, 18 - Taxes, 19 - Vacation, 
# 20 - Wedding Loans

ListingCategory.levels = 1:20
ListingCategory.labels = c(
  # "Not Available",
  "Debt Consolidation",
  "Home Improvement",
  "Business",
  "Personal Loan",
  "Student Use",
  "Auto",
  "Other",
  "Baby&Adoption",
  "Boat",
  "Cosmetic Procedure",
  "Engagement Ring",
  "Green Loans",
  "Household Expenses",
  "Large Purchases",
  "Medical/Dental",
  "Motorcycle",
  "RV",
  "Taxes",
  "Vacation",
  "Wedding Loans"
)

# split Q from Y and reverse to Y Q
# order by Y and Q
# reverse back to Q Y and join
# use later to get a chronologically ordered factor
loan.origination.quarter.levels <- levels(loans$LoanOriginationQuarter) %>%
  strsplit(' ') %>%
  lapply(rev) %>%
  lapply(paste, collapse = ' ') %>%
  unlist() %>%
  sort() %>%
  strsplit(' ') %>%
  lapply(rev) %>%
  lapply(paste, collapse = ' ') %>%
  unlist()

past.due.days.levels = c(
  "Past Due (1-15 days)",
  "Past Due (16-30 days)",
  "Past Due (31-60 days)",
  "Past Due (61-90 days)",
  "Past Due (91-120 days)",
  "Past Due (>120 days)"
)

loans <- loans %>% 
  mutate(
    ListingCreationDate = ListingCreationDate %>% 
      strptime('%Y-%m-%d %H:%M:%S') %>% 
      as.POSIXct(origin = lubridate::origin),
    ClosedDate = ClosedDate %>% 
      strptime('%Y-%m-%d %H:%M:%S') %>% 
      as.POSIXct(origin = lubridate::origin),
    DateCreditPulled = DateCreditPulled %>% 
      strptime('%Y-%m-%d %H:%M:%S') %>% 
      as.POSIXct(origin = lubridate::origin),
    LoanOriginationDate = LoanOriginationDate %>% 
      strptime('%Y-%m-%d %H:%M:%S') %>% 
      as.POSIXct(origin = lubridate::origin),
    FirstRecordedCreditLine = FirstRecordedCreditLine %>% 
      strptime('%Y-%m-%d %H:%M:%S') %>% 
      as.POSIXct(origin = lubridate::origin),
    IncomeRange = ordered(
      IncomeRange,
      levels = c( # ommit "Not employed", "Not displayed"
        "$0",
        "$1-24,999",
        "$25,000-49,999",
        "$50,000-74,999",
        "$75,000-99,999",
        "$100,000+"
      )
    ),
    ProsperRating..Alpha. = ordered(
      ProsperRating..Alpha.,
      levels = c("HR", "E", "D", "C", "B", "A", "AA")
    ),
    CreditGrade = ordered( # let's assume same order as ProsperRating
      CreditGrade,
      levels = c("NC", "HR", "E", "D", "C", "B", "A", "AA")
    ),
    LoanOriginationQuarter = ordered(
      LoanOriginationQuarter,
      levels = loan.origination.quarter.levels
    ),
    ListingCategory = factor(
      loans$ListingCategory..numeric.,
      levels = ListingCategory.levels,
      labels = ListingCategory.labels
    ),
    IncomeVerifiable = IncomeVerifiable == "True",
    CurrentlyInGroup = CurrentlyInGroup == "True",
    IsBorrowerHomeowner = IsBorrowerHomeowner == "True",
    # joined column for credit grade and prosper rating under assumption,
    # that same codes stand for same categories
    CreditGrade.ProsperRating = coalesce(CreditGrade, ProsperRating..Alpha.),
    # let's make ordeder factor for past due days
    past.due.days = ordered(LoanStatus, levels = past.due.days.levels)
  )
```


# Univariate Plots Section

```{r echo=FALSE, Structure}
str(loans)
```

```{r echo=FALSE, Summary}
summary(loans)
```

```{r echo=FALSE, Univariate_Plots}
# checkpoint
```

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = ListingCreationDate)
) +
  geom_histogram(binwidth = 60 * 60 * 24 * 30) +
  scale_x_datetime(date_breaks = "1 year", date_labels = "%Y")
```

There's a gap in late 2008 - early 2009. Must be related to some data points 
only having values until / from 2009.

Besides a slowdown in early 2013, number of listings increases almost 
exponentially since 2009.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = LoanOriginalAmount)
) +
  geom_histogram(binwidth = 1000) +
  scale_x_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000))
```

Round figures are very popular for original loan amounts.

```{r echo=FALSE}
ggplot(
  data = subset(loans, StatedMonthlyIncome <= 50000),
  aes(x = StatedMonthlyIncome)
) +
  geom_histogram() +
  scale_x_sqrt(breaks = seq(0, 50000, 2000)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Stated monthly income distribution is massively sqewed with some peope claiming 
to earn over a million per month.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = BorrowerRate)
) +
  geom_histogram(binwidth = 0.01)
```

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = BorrowerAPR)
) +
  geom_histogram(binwidth = 0.01)
```

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = LenderYield)
) +
  geom_histogram()
```

Borrower rate, APR and lender yield all have very similar values and follow
very similar distributions.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = DebtToIncomeRatio)
) +
  geom_histogram(binwidth = 0.1) +
  scale_x_sqrt(breaks = seq(0, 10, 0.5)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Debt to income ratio is capped at 10, giving a fat end for the tail of 
distribution, but majority of listings has much healthier ratios. Let's zoom in
a bit.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = DebtToIncomeRatio)
) +
  geom_histogram(binwidth = 0.05) +
  scale_x_continuous(breaks = seq(0, 1, 0.05)) +
  coord_cartesian(xlim = c(0, 1)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

A very smooth sqewed distribution here. Nothing unusual about it.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = as.character(Term))
) +
  geom_bar()
```

One year term loans are unexpectedly rare.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = Investors)
) +
  geom_histogram() +
  scale_x_continuous(breaks = seq(0, 1200, 100))
```

```{r echo=FALSE}
ggplot(
  data = subset(loans, Investors <= 30),
  aes(x = Investors)
) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 0:30)
```

All listings have at least 1 investor, but very few have more.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = ListingCategory)
) +
  geom_bar() +
  scale_y_sqrt(breaks = seq(0, 60000, 6000)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Debt consolitation is by far the most popular listing category. Counts between
categories vary greatly.

```{r echo=FALSE}
ggplot(
  data = subset(loans, Occupation != ""),
  aes(x = Occupation)
) +
  geom_bar() +
  scale_y_sqrt(breaks = seq(0, 30000, 3000)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Vast majority of occupations do not fall under predefined categories.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = IncomeRange)
) +
  geom_bar()
```

Besides some loan listings actually declaring $0 income, nothing surprising
in distribution of income range.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = PercentFunded)
) +
  geom_histogram(binwidth = 0.02) +
  scale_x_continuous(breaks = seq(0.6, 1.1, 0.02)) +
  scale_y_sqrt(breaks = seq(0, 120000, 6000))
```

There's some not fully funded listings. Distribution tail thickens going away 
from main bulk of values. This is curious.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = Recommendations)
) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 0:40) +
  scale_y_sqrt(breaks = seq(0, 120000, 6000))
```

Small number of extreme values. Having a recommendation is very rare.

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(past.due.days)),
  aes(x = past.due.days)
) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Small past due periods are more common than longer ones. Probably becasue this
catches some people that forgot to pay in time. Past 15 days, distribution stays
fairly flat.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = IsBorrowerHomeowner)
) +
  geom_bar()
```

A little over half listings are from home owners.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = CurrentlyInGroup)
) +
  geom_bar()
```

It's not typical to belong to a group.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = IncomeVerifiable)
) +
  geom_bar()
```

It's typical to have verifiable income.


# Univariate Analysis

### What is the structure of your dataset?

113937 observations and 81 data points, plus 3 calculated data points.

- # TODO describe

### What is/are the main feature(s) of interest in your dataset?

With so many data points the dataset could be split into multiple slices each
telling something important about it. I'm sure I've missed some important 
features, but of the ones I looked at these were most interesting: 
LoanOriginalAmount, Term, Investors, PercentFunded, Income, IsBorrowerHomeOwner.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

- past.due.days ordered factor based on LoanStatus to look more into past due
loans;
- CreditGrade.ProsperRating that is a joined version of ProsperRating..Alpha. 
and CreditGrade, because their levels seem to carry same meaning and a listing 
never has both present;
- ListingCategory based on ListingCategory..numeric. and description in the data
point definition to avoid having to memorize arbitrary IDs in order to perform
any kind of analysis about it.

### Of the features you investigated, were there any unusual distributions? \

- PercentFunded has a tail that thickens moving away from bulk of values;
- LoanOriginalAmount spikes at round figures.


# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots}
# checkpoint
```


```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = LoanOriginalAmount, y = Investors)
) +
  geom_point(alpha = 0.05) +
  scale_x_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000))
```

We see same clusters of observations on round numbers. Dots stack to vertical
bars, meaning, that most listings tend to fall on small number of 
LoanOriginalAmount values.

```{r echo=FALSE}
unique.LoanOriginalAmounts.summary <- c(
    length(unique(loans$LoanOriginalAmount)),
    length(loans$LoanOriginalAmount),
    length(unique(round(loans$LoanOriginalAmount / 500, digits = 0))),
    length(loans$LoanOriginalAmount[loans$LoanOriginalAmount %% 500 == 0])
  )
names(unique.LoanOriginalAmounts.summary) <- c(
  "unique.LoanOriginalAmounts",
  "all.listings",
  "unique.LoanOriginalAmounts.divisible.by.500",
  "listings.with.LoanOriginalAmounts.divisible.by.500"
)
unique.LoanOriginalAmounts.summary
```

Most loan amounts are divisible by 500.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(
    x = cut(
      LoanOriginalAmount,
      breaks = seq(250, max(loans$LoanOriginalAmount) + 250, 500),
      dig.lab = 10
    ),
    y = Investors
  )
) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(0, 1200, 100)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

```{r echo=FALSE}
ggplot(
  data = subset(loans, LoanOriginalAmount < 10250),
  aes(
    x = cut(
      LoanOriginalAmount,
      breaks = seq(250, 10250 + 250, 500),
      dig.lab = 10
    ),
    y = Investors
  )
) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(0, 600, 50)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Investor counts tends to increase as loan amount increases, but for most
ranges small investor counts are more likely.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = LoanOriginalAmount, y = LoanOriginalAmount / Investors)
) +
  geom_point(alpha = 0.05) +
  scale_x_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000)) +
  scale_y_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000))
```

Clear diagonal lines mark loans with small number of investors. We also have
a lot of loans where loan amount per investor is very small, indicating that
investing a small amount is popular acros all loan amounts.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = ListingCreationDate, y = LoanOriginalAmount)
) +
  geom_point(alpha = 0.05) +
  scale_y_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000)) +
  scale_x_datetime(date_breaks = "1 year", date_labels = "%Y")
```

We can see loans over $25000 only started appearing since 2013.

Loans under $2000 stopped appearing since 2011.

Decrease in loan count on early 2013 is visible across all amounts.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = ListingCreationDate, y = Investors)
) +
  geom_point(alpha = 0.05) +
  scale_y_continuous(breaks = seq(0, max(loans$Investors), 100)) +
  scale_x_datetime(date_breaks = "1 year", date_labels = "%Y")
```

Starting 2013 small number of inverstors per loan became much more popular.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = as.character(Term), y = LoanOriginalAmount)
) +
  geom_boxplot()
```

LoanOriginalAmount and Term relationship looks as expected. Longterm loans 
tend to be bigger.

```{r echo=FALSE}
ggplot(
  data = subset(loans, PercentFunded < 1),
  aes(x = LoanOriginalAmount, y = PercentFunded)
) +
  geom_point(alpha = 0.05) +
  scale_x_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000)) +
  scale_y_continuous(breaks = seq(0.7, 1, 0.05))
```

Not fully funded listings have a very currious relationship between 
LoanOriginalAmount and PercentFunded. Overal correlation is weak, but dots fall 
into diagonal lines.

Diagonal lines seem to arrive to round loan amount values as they approach full 
funding. These values probably are the amounts originally asked for.

```{r echo=FALSE}
ggplot(
  data = subset(loans, PercentFunded < 1),
  aes(x = LenderYield, y = PercentFunded)
) +
  geom_jitter(alpha = 0.05)
```

LenderYield and PercentFunded are unrelated.

```{r echo=FALSE}
ggplot(
  data = subset(loans, PercentFunded < 1),
  aes(x = Investors, y = PercentFunded)
) +
  geom_jitter(alpha = 0.05) +
  scale_x_continuous(breaks = seq(0, 1200, 100)) +
  scale_y_continuous(breaks = seq(0.7, 1, 0.05))
```

Investors and PercentFunded are not related either.

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = LoanOriginalAmount)
) +
  geom_density(aes(color = CreditGrade.ProsperRating)) +
  scale_color_brewer()
```

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = CreditGrade.ProsperRating, y = LoanOriginalAmount)
) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000))
```

CreditGrade.ProsperRating and LoanOriginalAmount are related. LoanOriginalAmount
increases until rating C and then flattens out.

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(ListingCategory)),
  aes(x = ListingCategory, y = LoanOriginalAmount)
) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Some listing categories tend to have much bigger loans than others. Debt 
Consolidation is surprisingly big. Baby&Adoption loans tend to be even bigger 
than Wedding Loans.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = BorrowerAPR, y = BorrowerRate)
) +
  geom_point(alpha = 0.05)
```

Not surprisingly BorrowerAPR and BorrowerRate have a very strong relationship.
What's unexpected is that it neatly falls into bunch of straight lines.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = IsBorrowerHomeowner, LoanOriginalAmount)
) +
  geom_boxplot()
```

Homeowners tend to get bigger loans.

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(ListingCategory)),
  aes(x = ListingCategory)
) +
  geom_bar(aes(fill = IsBorrowerHomeowner), position = "dodge") +
  scale_y_sqrt(breaks = seq(0, max(loans$LoanOriginalAmount), 5000)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Most listing categories have fairly similar popularity between homeowners and
non-homeowners. One exception is Home Improvement, with much more homeowners.
This makes perfect sense.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = CurrentlyInGroup, LoanOriginalAmount)
) +
  geom_boxplot()
```

Borrowers that are in groups tend to get smaller loans.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = IncomeVerifiable, LoanOriginalAmount)
) +
  geom_boxplot()
```

Borrowers with non-verifiable income tend to have a little smaller loans and is
probably first sub-group I've looked at that doesn't have loans over $25000.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = IncomeVerifiable, y = StatedMonthlyIncome)
) +
  geom_boxplot()
```

Now that's some seriously extreme outliers. Let's zoom in.

```{r echo=FALSE}
ggplot(
  data = loans,
  aes(x = IncomeVerifiable, y = StatedMonthlyIncome)
) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 15000))
```

Borrowers with verifiale income have tighter distribution of stated monthly 
income.

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(ListingCategory)),
  aes(x = ListingCategory, EstimatedReturn)
) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(ListingCategory)),
  aes(x = ListingCategory, EstimatedReturn)
) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0.05, 0.15)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Looks like EstimatedReturn has little to do with ListingCategory. More popular
categories have more outliers.

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = CreditGrade.ProsperRating, EstimatedReturn)
) +
  geom_boxplot()
```

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = CreditGrade.ProsperRating, EstimatedReturn)
) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0.03, 0.2))
```

Credit grade clearly has a relationship with estimated return. Better the
grade smaller and more nealty distributed the estimated return.


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Homeowners seem to get larger loans. They also seem to be getting more loans 
listed in home improvements, business, debt consolidation and childcare 
categories. That tells a little about social status of homeowners.

Even though most loans are funded by a single investor regardless of loan
amount, having large amount of investors (small contributions) seems to be
popular regardless of loan amount as well.

Loan amount tencds to be larger for meter credit ratings up to C and stays
similar for higher ratings. This could be due to an upper limit of loans amounts
visible in dataset.

### What was the strongest relationship you found?

BorrowerRate and BorroweAPR.


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}
# checkpoint
```

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = CreditGrade.ProsperRating, y = LoanOriginalAmount / Investors)
) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000))
```

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = CreditGrade.ProsperRating, y = LoanOriginalAmount / Investors)
) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(0, 8000, 1000)) +
  coord_cartesian(ylim = c(0, 8000))
```

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = CreditGrade.ProsperRating, y = LoanOriginalAmount / Investors)
) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(0, 700, 100)) +
  coord_cartesian(ylim = c(0, 700))
```

Even though loan amount per investor tends to be much greather for A, B and C
ratings, centers of distributions are similar across credit ratings.

```{r echo=FALSE}
ggplot(
  data = subset(loans, Investors <= quantile(Investors, 0.99)),
  aes(x = LoanOriginalAmount, y = PercentFunded)
) +
  geom_point(aes(colour = Investors)) +
  scale_colour_gradient(low = I("#000066"), high = I("#ffffff")) +
  scale_x_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000)) +
  scale_y_continuous(breaks = seq(0.7, 1, 0.05))
```

PercentFunded doesn't depend on Investor count.

```{r echo=FALSE}
ggplot(
  data = subset(
    loans,
    Investors <= quantile(Investors, 0.99)
    & !is.na(CreditGrade.ProsperRating)
  ),
  aes(x = CreditGrade.ProsperRating, y = LoanOriginalAmount)
) +
  geom_jitter(aes(colour = Investors)) +
  scale_colour_gradient()
```

Investor count tends to be higher for best credit scores.

```{r echo=FALSE}
ggplot(
  data = subset(
    loans,
    Investors <= quantile(Investors, 0.99)
    & !is.na(ListingCategory)
  ),
  aes(x = ListingCategory, y = LoanOriginalAmount)
) +
  geom_jitter(aes(colour = Investors)) +
  scale_color_gradient() +
  scale_y_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(ListingCategory)),
  aes(x = ListingCategory, y = LoanOriginalAmount / Investors)
) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(ListingCategory)),
  aes(x = ListingCategory, y = LoanOriginalAmount / Investors)
) +
  geom_boxplot() +
  scale_y_sqrt(breaks = seq(0, 5000, 500)) +
  coord_cartesian(ylim = c(0, 5000)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Throughout listing categories distribution centers of loan amount per investor 
look similar. Most categories have strongly skewed distributions.

Personal and student loans seem to have more investors per loan amount.

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = CreditGrade.ProsperRating, y = LoanOriginalAmount)
) +
  geom_boxplot(aes(colour = IsBorrowerHomeowner), position = "dodge")
```

Regardless of credit grade homeowners tend to get larger loans. Both groups
are capped at same amounts per credit grade though.

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = ListingCreationDate, y = LoanOriginalAmount)
) +
  geom_point(aes(color = CreditGrade.ProsperRating)) +
  scale_color_brewer() +
  scale_y_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000)) +
  scale_x_datetime(date_breaks = "1 year", date_labels = "%Y")
```

There are some visible caps for lower credit grades. Starting 2013 some were
increased.

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = CreditGrade.ProsperRating, EstimatedReturn)
) +
  geom_jitter(aes(color = LoanOriginalAmount)) +
  geom_boxplot(outlier.colour = NA, outlier.shape = NA, fill = NA, color = I("#dd0000"))
```

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = LoanOriginalAmount, y = EstimatedReturn)
) +
  geom_point(aes(color = CreditGrade.ProsperRating)) +
  scale_color_brewer() +
  scale_x_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000))
```

Lower credit grades tend to get smaller loans and more widely distributed 
estimated returns.

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = LoanOriginalAmount, y = EstimatedReturn)
) +
  geom_point(aes(color = BorrowerRate))
```

As expected, borrower rate and estimated return seem related. As not expected, 
negative estimated returns have a lot of high borrower rates in the mix.

Loan amount doesn't seem related to borrower rate or estimated return, but 
their variablity decreases as loan amount increases.

```{r echo=FALSE}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = LoanOriginalAmount, y = EstimatedReturn)
) +
  geom_density_2d(alpha = 0.5) +
  facet_wrap(~ CreditGrade.ProsperRating, ncol = 3)
```

```{r echo=FALSE}
ggplot(
  data = subset(
    loans,
    !is.na(CreditGrade.ProsperRating)
    & year(ListingCreationDate) >= 2009
  ),
  aes(x = LoanOriginalAmount, y = EstimatedReturn)
) +
  geom_point(aes(color = Term), size = 1) +
  facet_grid(year(ListingCreationDate) ~ CreditGrade.ProsperRating) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

This is useless being so tiny, but we're looking looking at 5 variables at once!

It's still good enough to reasert some previous observations that were difficult
to see in previous plots. Change of loan amount caps per credit grade over time
for example. We also see that negative estimated returns was a temporary thing
and starting 2011 disappeared together with a lot of estimated return 
extreme values.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Homeowners tend to get larger loans across credit grades. I was expecting this
effect to diminish for higher credit grades, but that was not the case.

Variability of data points like estimated return decreases as loan amount 
increases.


------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = LoanOriginalAmount)
) +
  geom_density(aes(color = CreditGrade.ProsperRating)) +
  scale_color_brewer("Credit grade") +
  scale_x_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000)) +
  labs(
    x = "Original loan amount",
    y = "Density",
    title = "Original loan amount density by credit grade"
  )
```

### Description One

Lower credit grade borrowers dominate loan amounts under 5000, but quickly
disappear as amount increases.

All densities spike on loan amounts divisible by 5000. 

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(
  data = subset(loans, !is.na(CreditGrade.ProsperRating)),
  aes(x = CreditGrade.ProsperRating, EstimatedReturn)
) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0.03, 0.2)) +
  labs(
    x = "Credit grade",
    y = "Estimated return",
    title = "Estimated return distribution by credit grade"
  )
```

### Description Two

Credit grade clearly has a relationship with estimated return. Better the
grade smaller and more nealty distributed the estimated return.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(
  data = subset(
    loans,
    !is.na(CreditGrade.ProsperRating) 
    & !is.na(EstimatedReturn)
  ),
  aes(x = LoanOriginalAmount, y = EstimatedReturn)
) +
  geom_point(aes(color = CreditGrade.ProsperRating)) +
  scale_color_brewer("Credit grade") +
  scale_x_continuous(breaks = seq(0, max(loans$LoanOriginalAmount), 5000)) +
  labs(
    x = "Original loan amount",
    y = "Estimated return",
    title = "Estimated return by original loan amount"
  )
```

### Description Three

Lower credit grades tend to get smaller loans and more widely distributed 
estimated returns.

Better credit grades concentrate at the lower end of estimated return with an
exception of a bunch of negative estimated returns for credit grade HR.


------

# Reflection

This was a very challenging dataset to explore. Main struggle was analysis 
paralysis inducing number of data points. It feels like my explorations barely
touched the surface.

It was interesting to see hints of lender policy changes in time series as well 
as tendency to borrow round amounts. These high density values created difficult
overplotting problems.

What was unexpected is how variability of many variables decreases as loan
amount and listing creation date increases. This shows that lender has stricter
rules for larger loans, as well as rules got stricter between 2010 and 2011 in
general.

It was a bit surprising to see such a visible relationship between credit grade 
and estimated return.

There's still plenty of data points to investigate in future work. Initially
when I started working on this dataset I expected to find clues on how to 
predict how much could be borrowed based on borrower properties, but it doesn't
seem to contain rejected loans, so that is hardly possible as people don't 
always borrow largest amount possible. This dataset would be good to build 
models though.
