Lesson 5
========================================================

```{r}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(tidyr)
```

```{r}
pf <- read.delim('pseudo_facebook.tsv')
```

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

```{r}
pf.fc_by_age_gender <- pf %>% 
  filter(!is.na(gender)) %>% 
  group_by(age, gender) %>% 
  summarise(
    mean_friend_count = mean(friend_count),
    median_friend_count = median(friend_count),
    n = n()
  ) %>% 
  ungroup() %>% 
  arrange(age)
```

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
ggplot(
  aes(x = gender, y = age),
  data = subset(pf, !is.na(gender))
) +
  geom_histogram()
```

***

### Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries}
ggplot(
  data = pf.fc_by_age_gender,
  aes(x = age, y = mean_friend_count)
) +
  geom_line(aes(color = gender)) +
  scale_x_continuous(breaks = seq(15, max(pf.fc_by_age_gender$age), 5))
```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:

***

### Reshaping Data
Notes:

```{r}
install.packages('reshape2')
library(reshape2)
```


***

```{r}
pf.fc_by_age_gender %>% 
  subset(
    select = c('gender', 'age', 'median_friend_count')
  ) %>% 
  spread(
    gender,
    median_friend_count
  ) %>%
  mutate(ratio = male / female) %>% 
  ggplot(
    aes(x = age, y = ratio)
  ) +
  geom_line() +
  geom_hline(yintercept = 1, linetype = 2, alpha = 0.5)
```

```{r}
pf.fc_by_age_gender %>% 
  dcast(age ~ gender, value.var = 'median_friend_count') %>%
  mutate(ratio = male / female) %>% 
  ggplot(
    aes(x = age, y = ratio)
  ) +
  geom_line() +
  geom_hline(yintercept = 1, alpha = 0.3, linetype = 2)
```

***

```{r}
pf$year_joined <- floor(2014 - pf$tenure / 365)
```

***

### Ratio Plot
Notes:

```{r Ratio Plot}

```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}

```

***

### Cut a Variable
Notes:

```{r Cut a Variable}
pf$year_joined.bucket <- ordered(
  cut(
    pf$year_joined,
    breaks = c(2004, 2009, 2011, 2012, 2014)
  ),
  levels = c('(2004,2009]', '(2009,2011]', '(2011,2012]', '(2012,2014]')
)
```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}
ggplot(
  aes(x = age, y = friend_count),
  data = subset(pf, !is.na(year_joined.bucket))
) +
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = median) +
  geom_line(linetype = 2, stat = 'summary', fun.y = mean)
  scale_x_continuous(breaks = seq(15, 110, 5))
```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}

```

***

### Friending Rate
Notes:

```{r Friending Rate}
with(subset(pf, tenure >= 1), summary(friend_count / tenure))
```

***

### Friendships Initiated
Notes:

What is the median friend rate?

What is the maximum friend rate?

```{r Friendships Initiated}
ggplot(
  data = subset(pf, tenure >= 1),
  aes(x = tenure, y = friendships_initiated / tenure)
) +
  # geom_line(stat = 'summary', fun.y = mean, aes(color = year_joined.bucket)) +
  geom_smooth(aes(color = year_joined.bucket))
```

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

```{r}
yo <- read.csv('yogurt.csv')
yo$id <- factor(yo$id)
head(yo)
```

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
ggplot(data = yo, aes(x = price)) +
  geom_histogram()
```

***

### Number of Purchases
Notes:

```{r Number of Purchases}
yo <- yo %>% 
  transform(all.purchases = strawberry + blueberry + pina.colada + plain + mixed.berry)
```

***

### Prices over Time
Notes:

```{r Prices over Time}
ggplot(
  data = yo,
  aes(x = time, y = price)
) +
  geom_jitter(alpha = 0.1)
```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

***

```{r}
set.seed(42)
ids <- sample(levels(yo$id), 16)
yo.sample <- subset(yo, id %in% ids)
```

```{r Looking at Sample of Households}
ggplot(
  data = yo.sample,
  aes(x = time, y = price)
) +
  facet_wrap( ~ id) +
  geom_line() +
  geom_point(aes(size = all.purchases), pch = 13)
```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:

***

```{r}
library(GGally)
```

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

